Closing the empirical cycle

### 2022-11-09 manuscript/manuscript.Rmd
Add supplemental tables

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1,2"
    corresponding: yes
    address: "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email: "c.j.vanlissa@uu.nl"
  - name: "Lukas Beinhauer"
    affiliation: "3"
  - name: "Susan Branje"
    affiliation: "4"
  - name: "Wim H. J. Meeus"
    affiliation: "4"

affiliation:
  - id: "1"
    institution: "Utrecht University, department of Methodology & Statistics"
  - id: "2"
    institution: "Open Science Community Utrecht"
  - id: "3"
    institution: "Helmut-Schmidt-Universität, Department of Methodology and Statistics for Psychology"
  - id: "4"
    institution: "Utrecht University, department of Youth and Family"

authornote: |
  This is a preprint paper, generated from Git Commit # `r gert::git_commit_id()`. CVL conceived of the study, conducted the final analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used, and is available under restricted access (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.

abstract: |
  | **Research Highlights**
  | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
  | • Parenting may be less relevant for emotion regulation development than previously thought, and negative parenting may be more impactful than positive parenting.
  | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
  | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
  | This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM-forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents based on their predictor scores, while maximizing the similarity of developmental trajectories within each group. Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories. The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors with parents and friends, and internalizing and externalizing behavior. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost.
A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
<span id = "abundantlit">There is already an abundance of empirical research on emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
Still, that review identified specific limitations of the existing literature: many studies are narrow in scope, considering only a few predictors and thus missing the relative importance compared to other predictors.</span>
This problem is exacerbated because the literature is fragmented across sub-disciplines, each of which examines mostly predictors within its scope.
And finally, the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
Specifically,
<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016] was used as a person-centered analysis technique,
to identify the most important early indicators (at age 13) of developmental trajectories
of emotion regulation across adolescence (age 14 through 18).
The resulting insights indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle [@croneUnderstandingAdolescencePeriod2012].</span>
Armed with this knowledge, future research and interventions can focus on the most important predictors.
Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010; @schaferEmotionRegulationStrategies2017; @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the development of cognitive capacities lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance might help motivate adolescents to engage in adaptive exploration:
Seeking out experiences that challenge their socio-emotional competences and stimulate growth.
At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
For example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

Longitudinal studies of emotion regulation development in adolescence provide preliminary empirical support for this notion of diverging trajectories: 
Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories

An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23. Seminal studies have applied "clustering" methods to identify youth with similar developmental trajectories14,24–26. Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors. Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29. The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30. -->

@@

<!-- SB: you have convincingly argued why machine learning in general is important, yet I miss the details related to ER. I would recommend to add a paragraph summarizing theories of ER and the factors they consider important. Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
The present study set out to obtain a relatively comprehensive understanding of the predictors of emotion regulation development across adolescence.
<!-- identify the most important early indicators of developmental trajectories -->
<!-- of  -->
The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
This involves an important challenges, as
the number of potentially relevant risk factors is too large for conventional methods.
At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
This challenge can be overcome by machine learning methods that perform variable selection.
We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```
<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which 
received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers conducted six annual measurement waves to collect questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more sensitive to misspecifications of growth curve models [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

```{r figclass, fig.cap="Latent class analysis of risk profiles."}
knitr::include_graphics("../classplot.pdf")
```

@@

[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the median, $+/- 1 SD$ of a standard normal distribution (i.e., the $16^{th}$ and $84^{th}$ percentile).
For convenience, these percentiles are labeled as $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
For categorical predictors, all unique values were used.

@@

```{r figvarimp, fig.cap="Predictors ranked in order of relative importance. m: mother, f: father, b: best friend, s: sibling."}
knitr::include_graphics("../pcomb.pdf")
```

@@

Examining the variable importance rankings,
several themes emerged.
Specifically, results indicated that several individual differences - personality and related constructs - were ranked among the most important predictors.
For example, three of the big five personality dimensions rank among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
The Behavioral Inhibition Scale (BIS, 2), which reflects motivation to avoid aversive outcomes, ranked highly - although the BAS (73), its counterpart which measures motivation to approach goal-oriented outcomes,
ranked very low.
One dimension of empathy - personal distress (9), the tendency to experience aversive emotions in response to the misfortunes of others,
also ranked highly, and substantially higher than other dimensions of empathy.
Finally, self-concept clarity (11) also ranked highly; this construct represents the internal consistency of self-referential thought, and has close theoretical ties to identity [@schwartzDailyDynamicsPersonal2011].

A second emerging theme is that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
For instance, adolescent-perceived interparental conflict (8),
adolescents' conflict engagement (escalation) with best friends (10),
mothers (13) and fathers (20),
frequency (14) and negative emotionality (28) of adolescents' conflict with both parents,
and withdrawal from conflicts with best friends (18) and mothers (19).

A third emerging theme is that parenting was relatively underrepresented among the most important predictors;
the highest-ranking parental predictor was balanced relatedness (3),
a construct that represents the extent to which parents support and respect the adolescent's autonomy.
It further appeared that negative and intrusive aspects of parenting were more predictive than other positive practices:
Prohibition and control of friendship choice (16),
perceived intrusiveness and meddling (25),
psychological control (27),
and supportive criticism (33).
Positively valenced dimensions of parent-adolescent relationship quality
like maternal (54) and paternal (55) support were ranked low,
and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.

Finally, an emerging theme is that internalizing- and externalizing symptoms
often associated with emotion regulation ranked moderately in terms of predictive importance.
Of these, parent-reported externalizing behavior (5) ranked highest, followed by adolescent-reported anxiety (15) and depression (17).
Adolescent-reported externalizing behavior ranked substantially lower (30).

It is further of note that some themes that are considered to be theoretically relevant ranked low.
For instance, indicators of maternal (86) and paternal (87) socio-economic status were ranked the lowest,
as were other family demographic indicators, like religion (81, 82).<!--SB: Why is this important? CJ: Religion is part of moral development, SES, and is directly related to emotion regulation: https://link.springer.com/chapter/10.1007/978-94-017-8950-9_13--> 
Aspects of adolescents' peer networks,
such as bullying (60, 84) and victimization (51, 62, 83),
and contact with delinquent peers (85) also ranked low.
The same applied to substance use;
the most commonly used substances of alcohol (78) and cigarettes (61) were ranked very low; drug use, which is less common, ranked a bit higher than this (37, but see direction of association).

@@

Other factors were associated with poorer emotion regulation (i.e., potential risk factors or diagnostic aids),
including big five neuroticism, 
behavioral inhibition system,
externalizing and internalizing (depression and anxiety) problem behavior,
frequency and intensity of conflict with both parents,
and use of the conflict resolutions styles engagement and withdrawal.
Intrusive parenting behaviors were also negative predictors, including psychological control, control of friendships, paternal power, and constructive criticism.
All empathy dimensions except perspective taking predicted greater emotion dysregulation,
as did different daily negative emotions,
parental stress and interparental conflict,
age and pubertal development,
and sex (boys reported better emotion regulation).

@@

<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEA < .07$, $CFI > .99$, $SRMR < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.</span>

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
  - name: "Lukas Beinhauer"
    affiliation: "2"
  - name: "Susan Branje"
    affiliation: "3"
  - name: "Wim H. J. Meeus"
    affiliation: "3"

affiliation:
  - id: "1"
    institution: "Tilburg University, department of Methodology & Statistics"
  - id: "2"
    institution: "Helmut-Schmidt-Universität, Department of Methodology and Statistics for Psychology"
  - id: "3"
    institution: "Utrecht University, department of Youth and Family"

authornote: |
  This is a preprint paper, generated from Git Commit # `r substring(gert::git_commit_id(), 1, 8)`. CVL conceived of the study, conducted the reported analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used, and are available under restricted access (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. We thank the participating children, parents and schools.

abstract: |
  | **Research Highlights**
  | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
  | • Specific parenting practices may be less relevant than previously thought, and negative parenting practices may be more impactful than positive ones.
  | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
  | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
  | [As 20% of adolescents develop emotion regulation difficulties, it is important to identify important early predictors thereof. Using the machine learning algorithm SEM-forests, we ranked the importance of (87) candidate variables assessed at age 13 in predicting quadratic latent trajectory models of emotion regulation development from age 14-18. Participants were 497 Dutch adolescents. Results indicated that the most important predictors were individual differences (e.g., in personality), aspects of relationship quality and conflict behaviors with parents and peers, and internalizing and externalizing problems. Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – and negative parenting practices ranked higher than positive ones. We discuss implications for theory and interventions, and present an open source risk assessment tool, ERRATA.]{#abstract}

@@

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, [as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost via their implications for mental health [@leeAdolescentMentalHealth2014].]{#cost}
A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
<span id = "abundantlit">There is already an abundance of empirical research on adolescent emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
Still, that review identified specific limitations of the existing literature:
many studies narrowly focus on a few predictors within a specific research area,
and thus fail to provide insight into these predictors' relative importance compared to other predictors, including those in other domains.
Relatedly,
the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].</span>

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
Specifically,
<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016], a person-centered analysis technique, was used
to identify the most important early indicators (at age 13) of developmental trajectories
of emotion regulation across adolescence (age 14 through 18).
The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle through adolescence [@croneUnderstandingAdolescencePeriod2012].</span>
These insights can guide development of interventions that focus on the most important predictors,
and direct the formation of overarching theory that adequately represents essential phenomena.

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010; @schaferEmotionRegulationStrategies2017; @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the development of cognitive capacities lags behind [see @caseyBrakingAcceleratingAdolescent2011; @vanlissaLongitudinalInterplayAffective2014].
This imbalance might help motivate adolescents to engage in adaptive exploration:
Seeking out experiences that challenge their socio-emotional competences and stimulate growth.
At the same time, this imbalance may render some adolescents vulnerable to emotion dysregulation if they are not yet able to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles [@leeAdolescentMentalHealth2014].
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].

Longitudinal studies of emotion regulation development in adolescence provide preliminary empirical support for this notion of diverging trajectories: 
Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
<!-- Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories - most have done  REF write something about other studies examining risk factors -->
An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23. Seminal studies have applied "clustering" methods to identify youth with similar developmental trajectories14,24–26. Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors. Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29. The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30. -->

@@

<!-- SB: you have convincingly argued why machine learning in general is important, yet I miss the details related to ER. I would recommend to add a paragraph summarizing theories of ER and the factors they consider important. Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
The present study set out to obtain a relatively comprehensive understanding of the predictors of emotion regulation development across adolescence.
<!-- identify the most important early indicators of developmental trajectories -->
<!-- of  -->
The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
This involves an important challenge, as
the number of potentially relevant risk factors is too large for conventional methods.
At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
This challenge can be overcome by machine learning methods that perform variable selection.
We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```
<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which 
received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers conducted six annual measurement waves to collect questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
[The sample demographic information is reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

```{r figclass, fig.cap="Latent class analysis of risk profiles."}
knitr::include_graphics("../classplot.png")
```

@@

[For continuous predictors, three values of $s$ were used - namely the percentiles, computed on the observed data, corresponding to the mean and $+/- 1 SD$ of a standard normal distribution (i.e., the $50^{th}$, $16^{th}$ and $84^{th}$ percentiles).
For convenience, these percentiles are labeled as *mean* and $+/- 1 SD$ in Figure \@ref(fig:figpdp1)-\@ref(fig:figpdp4).]{#percentiles}
For categorical predictors, all unique values were used.

@@

```{r figvarimp, fig.cap="Predictors ranked in order of relative importance. m: mother, f: father, b: best friend, s: sibling."}
#readRDS("../varimp_comb2_03-08-2022.RData")
knitr::include_graphics("../pcomb2.png")
```

@@

Examining the variable importance rankings,
several themes emerged (see Figure \@ref(fig:figthemes) panel A).
Specifically, results indicated that among the most important predictors were personality and personality-related constructs.
Three of the big five personality dimensions ranked among the top six predictors: Neuroticism (1), extraversion (4), and agreeableness (6).
The Behavioral Inhibition Scale (BIS, 2), which reflects motivation to avoid aversive outcomes, ranked highly - although the BAS (73), its counterpart which measures motivation to approach goal-oriented outcomes,
ranked very low.
One dimension of empathy - personal distress (9), the tendency to experience aversive emotions in response to the misfortunes of others,
also ranked highly, and substantially higher than other dimensions of empathy.
Finally, self-concept clarity (11) also ranked highly; this construct represents the internal consistency of self-referential thought, and has close theoretical ties to identity [@schwartzDailyDynamicsPersonal2011].

```{r figthemes, fig.cap="Variable importances clustered by theme (A) and level of the bioecological model (B)."}
knitr::include_graphics("../violin.png")
```

A second emerging theme was that most constructs related to conflict frequency and behavior ranked highly, and few ranked low.
For instance, adolescent-perceived interparental conflict (8),
adolescents' conflict engagement (escalation) with best friends (10),
mothers (13) and fathers (20),
frequency (14) and negative emotionality (28) of adolescents' conflict with both parents,
and withdrawal from conflicts with best friends (18) and mothers (19).

<span id = "parenting">A third emerging theme is that specific parenting practices were relatively underrepresented among the most important predictors.
Parenting practices are defined as specific parental actions related to child-rearing,
for example, support and dimensions of (behavioral and psychological) control [@petersonFamilySocialCompetence1990].</span>
Negative and intrusive parenting practices were more predictive than positive ones:
Of these, prohibition and control of friendship choice (16) ranked highest, followed by intrusiveness and meddling (25),
psychological control (27),
and supportive criticism (33).
Positive parenting practices,
like maternal (54) and paternal (55) support,
were ranked low,
and neutrally valenced aspects like parental monitoring (sollicitation 71, adolescent disclosure 72) were ranked very low.
This is interesting, given the extensive research attention into the effect of these specific parenting practices.

<span id = "parentrelated">Several parent-related constructs scored higher than the aforementioned child rearing behaviors;
the highest of these was balanced relatedness (3),
which represents parents' tolerance for adolescents' differing opinions and is related to parent-child intimacy and autonomy [@shuhnanAdolescentIntimacyRevisited1997].
Other parent-related constructs that scored relatively highly were interparental conflict, which had a negative effect on emotion regulation (8);
like balanced relatedness, conflict likely contributes to the emotional family climate [@morrisRoleFamilyContext2007].
Finally, maternal age (7) had a protective effect, which is well-supported by prior literature [@zondervan-zwijnenburgParentalAgeOffspring2020].]</span>
<!-- Add reference -->

Finally, an emerging theme was that internalizing- and externalizing symptoms
often associated with emotion regulation ranked relatively highly in terms of predictive importance.
Of these, parent-reported externalizing behavior (5) ranked highest, followed by adolescent-reported anxiety (15) and depression (17).
Adolescent-reported externalizing behavior ranked substantially lower (30).

It is further of note that some themes that are considered to be theoretically relevant ranked low.
This was the case, for example, for demographic variables like maternal (86) and paternal (87) socio-economic status and family religion (81, 82).<!--SB: Why is this important? CJ: Religion is part of moral development, SES, and is directly related to emotion regulation: https://link.springer.com/chapter/10.1007/978-94-017-8950-9_13--> 
Aspects of adolescents' peer networks,
such as bullying (60, 84) and victimization (51, 62, 83),
and contact with delinquent peers (85) also ranked unexpectedly low.
The same applied to substance use;
the most commonly used substances of alcohol (78) and cigarettes (61) were ranked very low; drug use, which is less common, ranked a bit higher than this (37, but see direction of association).

@@

Other factors were associated with poorer emotion regulation (i.e., potential risk factors or diagnostic aids),
including big five neuroticism, 
behavioral inhibition system,
externalizing and internalizing (depression and anxiety) problem behavior,
frequency and intensity of conflict with both parents,
and use of the conflict resolutions styles engagement and withdrawal.
Intrusive parenting practices were also negative predictors, including psychological control, control of friendships, paternal power, and constructive criticism.
All empathy dimensions except perspective taking predicted greater emotion dysregulation,
as did different daily negative emotions,
parental stress and interparental conflict,
age and pubertal development,
and sex (boys reported better emotion regulation).

@@

<span id = "posthoc">A Reviewer suggested following up the planned machine learning analysis with follow-up analyses that might be more familiar to a developmental readership:
Regressing the parameters (intercept, slope and quadratic slope) of a latent growth model on each predictor in turn.
Such analyses might be helpful to plan hypotheses and perform power analysis for follow-up research within the SEM framework.
However, we strongly caution against over-interpretation.
Unlike the machine learning analyses, these analyses do not accommodate non-linear effects or interactions, nor are effects controlled for other predictors.
The results are merely descriptive and should not be used to perform inference.
[Supplemental Table S3](https://github.com/cjvanlissa/veni_forest/blob/master/supplemental_table_3.csv) reports standardized regression coefficients with significance asterisks for all continuous and binary predictors, and significance asterisks of Likelihood Ratio tests for multi-group categorical predictors.
All models had good fit ($RMSEAs < .07$, $CFIs > .99$, $SRMRs < .04$).
Out of all predictors, 79% had a significant effect on the intercept, 24% on the slope, and 18% on the quadratic slope.
This indicates that our machine learning analyses indeed recovered relevant patterns in the data.</span>
